A Vignetted View of Computer Science through Citation Analysis

### 2019-12-17 Scientometrics/comp.tex
added more text

\title{Computer Science, Vignettes, and Citation Analysis}
%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
\subtitle{Salton and Bergmark Redux}
%\titlerunning{Short form of title}        % if too long for running head

@@

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus, the Web of Science research and categorical classifications from Clarivate Analytics, and the NSF classification system~\cite{nsf_classification} [add citations for Scopus and Web of Science]. Scopus and the Web of Science also include comprehensive bibliographies with citation links and proprietary unique identifiers for publications. 
All three systems rely on applying one or more journal-derived labels to articles.  
A logical prediction, given the diversity of articles within journals, is limited specificity at the article level. 
Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. 
Others have also discussed and critiqued disciplinary assignments using journal-based classifications~\cite{wang_large-scale_2016,perianes-rodriguez_comparison_2017}. 
Article classification systems have been constructed that escape some of the criticisms of journal-based classification~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012}, but seem to be in the research domain rather than the domain of practical use [need to check on this].[May need another bridging sentence to the next paragraph].

In this study, we combine article and journal level approaches to extend the work of Salton and Bergmark to present trends in computer science, while considering connections to other fields. As a source of computer science literature, we used DBLP, a reference bibliography for computer science~\cite{dblp_ref}. The DBLP bibliography covers publications from computer science and includes publications from hybrid fields, where they are considered pertinent to computer science research.

We describe below the high-level interactions of computer science within the Physical Sciences, and with the Social Sciences, Life Sciences, and Health Sciences [What did we find? Some kind of preview sentence!].

@@

For the purpose of this study, our working definition of the computer science literature was all publications in the DBLP bibliography that (i) had a digital object identifier (DOI), and (ii) could also be matched to article identifiers in the Scopus bibliography. 
Our underlying assumption is that records in DBLP are greatly enriched for computer science even while recognizing that coverage may be incomplete and possibly biased. Cross-matching DBLP publications to records in the Scopus, abstract and citation database of peer-reviewed literature, enabled us to harvest the richer links in Scopus as well as extract links to publications from other disciplines. This cross-matching also allows the use of both journal and article based classifications when clustering documents. Using both direct citations and co-citations as the basis for clustering, we used DBLP articles from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the All Science Journal Classification (ASJC) developed and maintained by Scopus through a combination of automated and manual procedures, producing a dataset of  2,685,356 publications.

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Just over 95\% of the publications within were published after 1996. Publications were parsed from the XML source file and loaded into a PostgreSQL database.  We have previously parsed the Scopus dataset of over 72,000,000 publications into a custom schema as part of implementing a larger data platform for research evaluation~\cite{GithubERNIE2019}. The total number of publications in Scopus that are labeled COMP with major subject area of Computer Science is 5,835,160. Records in the DBLP dataset were matched to Scopus identifiers using digital object identifiers (DOIs). 
 This procedure resulted in a dataset of 2,685,356 DBLP publications with Scopus identifiers where 1,278,322 (47.6\%) were labeled as article and 1,407,034 (52.4\%) as conference proceedings.  
 References cited by these publications were then extracted from Scopus (7,129,006 records), resulting in a total of 8,000,411 nodes (publications and references) and 44,296,381 edges that represent citations within the dataset. 
 This dataset is referred to as \emph{comp} (Fig~\ref{fig:ar_cp_annotation}, Table~\ref{tab:comp}).

@@

We also used conductance, as defined in Shun et al.~\cite{shun_parallel_2016}, to  evaluate clustering (smaller is better), noting that conductance has been found to be ``the stand-alone quality metric that best indicates performance on information recovery metrics" \cite{emmons2016analysis}. 
%Tandy, George - add reference to Zaki paper as well
In our analysis, we saw that the last cluster (cluster 19) had a much larger conductance value than the other clusters and also had the smallest
number of nodes.
We then examined
results obtained using Graclus with two other numbers of clusters (18 and 22), and
in each case, the highest numbered cluster  had the greatest conductance value and also the smallest number of nodes.
These results suggest that Graclus produces a final cluster that effectively serves as a container for `left over publications' during the clustering procedure. 
Therefore, we limited our consideration of cluster 19 (the final of the twenty clusters)  when interpreting results.  The remaining 19 clusters had conductance values ranging from 0.09 to 0.25 with a median conductance of 0.15 (Fig.~\ref{fig:graclus_comparison_conductance}, Table~\ref{tab:comp}).

\paragraph{Clustering by Co-Citation.}
\textcolor{blue}{Check the next paragraphs for repetitiveness.}

For an alternate view of these DBLP data we constructed clusters using co-citation, the frequency with which a pair of articles is cited by other articles. 
Unlike clustering by direct citation with Graclus, where every input publication is assigned to a cluster and every citation is weighted equally, clustering by co-citation is progressively selective, involves assigning weights to co-citations, and also accounts for weak inter-cluster interactions.

Co-citation was first described independently by Small and Marshakova in 1973~\cite{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}, and provides insight into the emergence of new ideas derived from the association of previously independent ones.  Co-citation thus provides some degree of evolutionary insight and has been used for clustering publications~\cite{boyack_cocitation_2010,boyack_improving_2013,small_structure_1974,small_clustering_1985}.

We first identified the top 10\% most-cited articles in the \emph{comp} dataset and then documented all co-cited pairs of their cited references that were also elements of \emph{comp}. The frequency of co-citation was then computed across all records in Scopus and normalized using Salton's correction. These co-cited pairs and the normalized co-citation frequencies were then represented as a graph where each node was a publication and each edge represented a co-citation, weighted by the normalized co-citation frequency.

@@

Agglomerative clustering was then performed on these 10,298 clusters to generate higher-order clusters. To focus on larger clusters, only those with at least 10 nodes were used as input. Briefly, each cluster was now treated as a node and the edge weight between two clusters was assigned to the maximum edge weight of all edges between the nodes in the two precursor clusters. Edges were arranged in descending order. The first pair of clusters was merged and its edge weight with other interacting clusters was recalculated, again based on maximum edge weight. All edges were then re-ordered as before and the next pair of clusters was merged. The process was halted after 600 rounds to prevent a single very large outlier cluster being generated (Fig.~\ref{fig:merge_select}).

@@

\subsection{Data}
The process of selection and matching resulted in a dataset of 2,685,356 publications (Materials and Methods). 
Of these roughly 2.8 million publications, approximately 2.07 million were classified as COMP under subject area and assigned ASJC codes corresponding to Computer Science (major subject area with 13 minor subject areas (Table~\ref{tab:comp}), with the balance of 610,000 publications non-exclusively shared between 26 different major subject areas ranging from Engineering (330,048) to Dentistry (Fig.~\ref{fig:ar_cp_annotation}).

The set of 2.07 million publications classified as COMP in Scopus spanned all 13 minor subject areas with Software 30.3\% being the largest component and Computer Science (miscellaneous) 0.9\% the smallest (Table~\ref{tab:comp}). The number of minor subject areas assigned to a publication averaged to 1.75 with 53\% of the publications being assigned to one minor subject area and 0.0003\% assigned to 7 different minor subject areas. 
We note also that  the $4.8x10^6$ DBLP publications are only partially found in Scopus (which had a total of $5.8x10^6$ publications), using our methodology.
%George, please consider revising this text
%the sets of computer science publications in DBLP and Scopus are partially overlapping rather than DBLP being a perfect subset of Scopus. [the point being that we should justify our use of both journal and article level classifications].

@@

\subsection{Clustering by Co-citation.} For an alternate examination of the data, we used co-citation frequencies to cluster the DBLP dataset. 
Figure \ref{fig:heatmap_cocit} shows a heatmap in which clusters constructed by co-citation are mapped to Scopus ASJC minor subject areas. Mappings are only shown if they account for at least 15\% of the publications in a cluster. 
\textcolor{blue}{
Unlike the earlier figure [insert figure reference] Unlike the case of clustering by direct citation, no cluster maps to more than three ASJC minor subject areas. We have to decide what to say about this figure but I think it suggests intra-cluster relationships for the most part. }

@@

\begin{table}[ht]
\caption{\emph{Statistics regarding the 20 clusters produced using direct citation. The \emph{comp} dataset 
2685021 is because of inner joins on ASJC Explain conductance calculations~\cite{shun_parallel_2016}}}
\label{tab:comp}       
\begin{tabular}{lrccc}
\hline\noalign{\smallskip}
minor\_subject\_area & Percent of Publications \\
\noalign{\smallskip}\hline\noalign{\smallskip}
Software & 30.30 \\ 
Computer Science Applications & 25.20 \\ 
Computer Networks and Communications & 21.60 \\ 
Computer Science(all) & 19.60 \\ 
Artificial Intelligence & 14.30 \\ 
Information Systems & 11.90 \\ 
Hardware and Architecture & 11.50 \\ 
Signal Processing & 9.90 \\ 
Computer Vision and Pattern Recognition & 9.50 \\ 
Computational Theory and Mathematics & 8.00 \\ 
Human-Computer Interaction & 7.50 \\ 
Computer Graphics and Computer-Aided Design & 5.20 \\ 
Computer Science (miscellaneous) & 0.90 \\ 
\noalign{\smallskip}\hline
\end{tabular}
\end{table}
\newpage

>>>>

\title{A Vignetted View of Computer Science through Citation Analysis}
%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
\subtitle{Salton and Bergmark Redux}
%\titlerunning{Short form of title}        % if too long for running head

@@

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus, the Web of Science research and categorical classifications from Clarivate Analytics, and the NSF classification system~\cite{nsf_classification} [add citations for Scopus and Web of Science]. Scopus and the Web of Science also include comprehensive bibliographies with citation links and proprietary unique identifiers for publications. 
All three systems rely on applying one or more journal-derived labels to articles.  
A logical prediction, given the diversity of articles within journals, is limited specificity at the article level. 
Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. 
Others have also discussed and critiqued disciplinary assignments using journal-based classifications~\cite{wang_large-scale_2016,perianes-rodriguez_comparison_2017}. 
Article classification systems have been constructed that escape some of the criticisms of journal-based classification~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012}, but do not enjoy widespread use. [Add reference to Scandinavian paper in QSS].

In this study, we extend the work of Salton and Bergmark by combining article and journal approaches to study trends in computer science, while also considering connections to other fields. As a source of computer science literature, we used DBLP, a reference bibliography for computer science~\cite{dblp_ref}. The DBLP bibliography covers publications from computer science and includes publications from hybrid fields, where they are considered pertinent to computer science research. We describe below the high-level interactions of computer science within the Physical Sciences, and with the Social Sciences, Life Sciences, and Health Sciences [What did we find? Some kind of preview sentence!].

@@

For the purpose of this study, our working definition of the computer science literature was all publications in the DBLP bibliography that (i) had a digital object identifier (DOI), and (ii) could also be matched to article identifiers in the Scopus bibliography. 
Our underlying assumption is that records in DBLP are greatly enriched for computer science even while recognizing that coverage may be incomplete and possibly biased. Cross-matching DBLP publications to records in the Scopus, abstract and citation database of peer-reviewed literature, enabled us to harvest the richer links in Scopus as well as extract links to publications from other disciplines. This cross-matching to Scopus, also allows the use of both journal and article based classifications when clustering documents. Using both direct citations and co-citations as the basis for clustering, we used DBLP articles from journals and conference proceedings for the period 1996-2015. We reconciled these clusters with the All Science Journal Classification (ASJC) developed and maintained by Scopus through a combination of automated and manual procedures, producing a dataset of  2,685,356 publications.

\subsection{Data}
A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Just over 95\% of the publications within were published after 1996. Publications were parsed from the XML source file and loaded into a PostgreSQL database.  We have previously parsed the Scopus dataset, presently at over 87.8 million publications, into a custom schema as part of implementing a larger data platform for research evaluation~\cite{GithubERNIE2019}. The total number of publications in Scopus that are labeled COMP with major subject area of Computer Science is 5,835,160. Records in the DBLP dataset were matched to Scopus identifiers using digital object identifiers (DOIs). This procedure resulted in a dataset of 2,685,356 DBLP publications with Scopus identifiers where 1,278,322 (47.6\%) were labeled as article and 1,407,034 (52.4\%) as conference proceedings.  References cited by these publications were then extracted from Scopus (7,129,006 records), resulting in a total of 8,000,411 nodes (publications and references) and 44,296,381 edges that represent citations within the dataset. This dataset is referred to as \emph{comp} (Fig~\ref{fig:ar_cp_annotation}, Table~\ref{tab:comp}).

@@

We also used conductance, as defined in Shun et al.~\cite{shun_parallel_2016}, to  evaluate clustering by direct citation (smaller is better), noting that conductance has been found to be a good metric for this purpose~\cite{emmons2016analysis,almeida_2012}. 
%Tandy, George - add reference to Zaki paper as well
In our analysis, we saw that the last cluster (cluster 19) had a much larger conductance value than the other clusters and also had the smallest
number of nodes.
We then examined
results obtained using Graclus with two other numbers of clusters (18 and 22), and
in each case, the highest numbered cluster  had the greatest conductance value and also the smallest number of nodes.
These results suggest that Graclus produces a final cluster that effectively serves as a container for `left over publications' during the clustering procedure. 
Therefore, we limited our consideration of cluster 19 (the final of the twenty clusters)  when interpreting results.  The remaining 19 clusters had conductance values ranging from 0.09 to 0.25 with a median conductance of 0.15 (Fig.~\ref{fig:graclus_comparison_conductance}, Table~\ref{tab:comp}).

\paragraph{Clustering by Co-Citation.}

For an alternate view of these DBLP data we constructed clusters using co-citation, the frequency with which a pair of articles is cited by other articles\cite{small_co-citation_1973,marshakova-shaikevich_co-citation_1973} and has been used for clustering publications. Co-citation, first described independently by Small and Marshakova in 1973, provides insight into the emergence of new ideas derived from the association of previously independent ones. Unlike clustering by direct citation with Graclus, where every input publication is assigned to a cluster and every citation is weighted equally, clustering by co-citation is progressively selective, involves assigning weights to co-citations, and also accounts for weak inter-cluster interactions~\cite{boyack_cocitation_2010,boyack_improving_2013,small_structure_1974,small_clustering_1985}.

@@

Agglomerative clustering was then performed on these 10,298 clusters to generate higher-order clusters. To focus on larger clusters, only those with at least 10 nodes were used as input. Briefly, each cluster was now treated as a node and the edge weight between two clusters was assigned to the maximum edge weight of all edges between the nodes in the two clusters. Edges were arranged in descending order. The first pair of clusters was merged and its edge weight with other interacting clusters was recalculated, again based on maximum edge weight. All edges were then re-ordered as before and the next pair of clusters was merged. The process was halted after 600 rounds to prevent large outlier clusters being generated (Fig.~\ref{fig:merge_select}).

@@

\subsection{Data}
The process of selection and matching resulted in a dataset of  publications (Materials and Methods). Of 4,291,130 DBLP publications with DOIs, only  2,685,356 are found in Scopus. Of these roughly 2.8 million publications, approximately 2.07 million were classified in Scopus as COMP under subject area and assigned ASJC codes corresponding to Computer Science (major subject area with 13 minor subject areas (Table~\ref{tab:comp}), with the balance of 610,000 publications non-exclusively shared between 26 different major subject areas ranging from Engineering (330,048) to Dentistry (Fig.~\ref{fig:ar_cp_annotation}). The set of 2.07 million publications classified under the major subject area Computer Science in Scopus spanned all 13 minor subject areas with Software 30.3\% being the largest component and Computer Science (miscellaneous) 0.9\% the smallest (Table~\ref{tab:comp}). Another 409,082  publications were labeled Theoretical Computer Science under the major subject area of Mathematics. 38.86\% of the publications were assigned to one minor subject area, 78.1\% to two minor subject areas, 909.9\% to three minor subject areas, and 0.0003\% assigned to 8 different minor subject areas.

%George, please consider revising this text
%the sets of computer science publications in DBLP and Scopus are partially overlapping rather than DBLP being a perfect subset of Scopus. [the point being that we should justify our use of both journal and article level classifications].

@@

\subsection{Clustering by Co-citation.} For an alternate examination of the data, we used co-citation frequencies to cluster the DBLP dataset as described above. Figure \ref{fig:heatmap_cocit} shows a heatmap in which clusters constructed by co-citation are mapped to Scopus ASJC minor subject areas. Mappings are only shown if they account for at least 15\% of the publications in a cluster (upper panel). Under these conditions, only 17 of 20 clusters mapped to minor subject areas where at least 15\% of the nodes in a cluster were labeled with a given subject area. When the threshold was reduced to 10\% all 20 clusters indicated at least one mapping to a minor subject area. We interpret these data as indicative of two kinds of co-citational linkages (i) intra-cluster (ii) cross cluster. The latter predominates and is detected at lower thresholds (lower panel). No cluster maps to more than 3 minor subject area at the 15\% threshold in comparison to a maximum of 8 minor subject areas for clustering by direct citation (Fig 3, Cluster 17). It should be noted, however, that the size of clusters constructed by co-citation is much smaller than those generated by direct citation through Graclus.

\textcolor{blue}{
Unlike the earlier figure [insert figure reference] Unlike the case of clustering by direct citation, no cluster maps to more than three ASJC minor subject areas. We have to decide what to say about this figure but I think it suggests intra-cluster relationships for the most part. }

@@

\begin{table}[ht]
\caption{\emph{denominator is 2090947 (some source ids got lost via inner joins and code restrictions) Statistics regarding the 20 clusters produced using direct citation. The \emph{comp} dataset 
2685021 is because of inner joins on ASJC Explain conductance calculations~\cite{shun_parallel_2016}}}
\label{tab:comp}       
\begin{tabular}{lrccc}
\hline\noalign{\smallskip}
minor\_subject\_area & Percent of Publications \\
\noalign{\smallskip}\hline\noalign{\smallskip}
Software & 30.1 \\
Computer Science Applications & 25.0 \\ 
Computer Networks and Communications & 21.5 \\
Theoretical Computer Science & 19.6 \\ 
Computer Science(all) & 19.4 \\ 
Artificial Intelligence & 14.2 \\ 
Information Systems & 11.8 \\ 
Hardware and Architecture & 11.8 \\ 
Signal Processing & 9.8 \\ 
Computer Vision and Pattern Recognition & 9.4 \\ 
Computational Theory and Mathematics & 7.9 \\ 
Human-Computer Interaction & 7.4 \\ 
Computer Graphics and Computer-Aided Design & 5.2 \\ 
Computer Science (miscellaneous) & 0.9 \\ 
\noalign{\smallskip}\hline
\end{tabular}
\end{table}
\newpage
